Looking for a Signal in the Noise: Revisiting Obesity and the microbiome

### 2016-03-25 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
manuscript update writing

@@

**Results:** When we considered each dataset in isolation, there were no significant differences (P > 0.05) in B/F between non-obese and obese individuals. Only one diversity measure (OTU richness (N=1)) was significantly different between non-obese and obese individuals (P<0.05). With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals (mean AUC: `r format(round(mean(AUCRFData$OOBAUCAll), 2), nsmall=2)`; s.d.=`r format(round(sd(AUCRFData$OOBAUCAll), 2), nsmall=2)`). In 7/8 of the studies an OTU related to the Ruminococcaceae family was important in classification; however, the direction of the association varied between studies. When we pooled the results across studies, there was a significant difference in RR of obesity based on Shannon diversity (RR = 1.16, CI = 1.01-1.34, P-value = 0.04) but not B/F ratio (RR = 0.98, CI = 0.85-1.13, P-value = 0.81). Finally, when we normalized the Shannon diversity data and pooled studies there was a significant difference between non-obese and obese individuals (difference in mean z-score = `r format(round(combZHTtest$estimate[1]-combZHTtest$estimate[2], 2), nsmall=2)`; P = `r format(round(combZHTtest$p.value, 2), nsmall=2)`). Similar to RR there was no significant difference in the normalized B/F ratio between non-obese and obese individuals (difference in mean z-score = `r format(round(combZLogBFTtest$estimate[1]-combZLogBFTtest$estimate[2], 2), nsmall=2)`; P = `r format(round(combZLogBFTtest$p.value, 2), nsmall=2)`).

**Conclusion:** We show that by pooling the results of many individual studies we can test the generalizability of the microbiome. Our meta-analysis supports the  hypothesis that there is a difference between non-obese and obese individuals in Shannon diversity and B/F ratio, but that the effect sizes are small.

@@

***Data Analysis***
We performed the analysis using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`). We split the overall analysis into three general strategies. In the first strategy we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B/F), Shannon diversity, and number of OTUs. A two-tailed Wilcoxon test was performed for comparison for non-obese and obese individuals. We also measured the Spearman correlation coefficient between each metric and the actual BMI value. Next, we compared the community structure from non-obese and obese individuals using PERMANOVA analysis of Bray-Curtis distance matrices. This analysis was performed using the vegan R package. Finally, we used the AUCRF R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguish non-obese and obese individuals.

@@

In the third strategy, we normalized the diversity and B/F data within each study and pooled the normalized data across studies. We applied a Z-score normalization to the Shannon diversity data and log-transformed B/F data within each study to have a mean of zero and standard deviation of one. We then pooled Z-scores across studies and used a two-tailed T-test on the values to test for a significant difference in Z-scores among non-obese and obese individuals. To assess differences between the lean, overweight, and obese BMI categories we used ANOVA with a Tukey post hoc test. We also measured the Spearman correlation coefficient between the Z-scores for each metric and BMI values.

@@

***Separate Analysis***
We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicute (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014]. Using a P-value cutoff for significance as less than 0.05 we found a total of 1 significant result for Shannon diversity, 2 significant results for OTU richness, and 1 signficant result for the Bray-Curtis distance matrix. There were no significant results for Bacteroidetes, Firmicutes, or the B/F ratio. The full summary of the exact P-values for each respective study and variable tested can be found in table 3.

Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm. We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(AUCRFData$OOBAUCAll), 2), nsmall=2)` +/- `r format(round(sd(AUCRFData$OOBAUCAll), 2), nsmall=2)`. We found that the total range for the total number of variables used for the classification ranged from 3 (Ross and Escobar) to 40 (Zupancic). The total number of trees and node size was standardized to 1000 and 20 respectively for each data set. One aspect we found surprising was that there was very little overlap in the variables used by the random forest algorithm for the classification of obese and non-obese individuals by study. The three most common variables was for an OTU that could be classified to the family Ruminococcaceae (7/8 studies), an OTU that could be classified to the family Lachnospiraceae (6/8 studies), Clostridiales (6/8) and Shannon diversity metric (5/8 studies).

@@

We next moved on to the pooled meta-analysis. For the Shannon diversity RR, all studies but Wu et al had those with a lower Shannon diversity then the median as being higher risk for obesity [Figure 2]. This was confirmed with the pooled analysis showing an increased RR of 1.16 (CI 1.01, 1.34) for obesity in the lower then median Shannon diversity group (P-value = 0.0412). When investigated the B/F ratio RR most studies showed that there was no significant difference between obese and non-obese individuals [Figure 3].  The pooled RR was 0.98 (CI 0.85, 1.13) suggesting that a low B/F ratio does indeed have less obese individuals (P-value = 0.811).

We also used a funnel plot to asses the bias of the 9 data sets used in this meta-analysis. From the graph we can observe that there does not seem to be any significant  bias [Figure 4]. Both funnel plots have almost equal number of studies scattered on either side of the estimated true value.

The final forest plot that we used for this analysis investigated how the different OTUs, classified to Ruminococcaceae, that were important for the classification of obese and non-obese based on the OOB AUC in the the separate study analysis looked when pooled together to assess the RR [Figure S4]. There was no significant increase in RR based on the Ruminococcaceae OTU (RR = 0.96 (CI 0.60, 1.52), P-value = 0.8575). Thus it would seem that although Ruminococcaceae is important for classification the directionality is study dependent [Figure S4].

@@

Using the third approach with Z-score normalization we found that there was a significant difference in the Shannon diversity between the obese and non-obese group (P-value = 0.00493) [Figure 5A]. Further there was also a significant difference in this normalized Shannon diversity value when the data was split between the three BMI groups (P-value = 0.03826) [Figure 5B]. Using a Tukey post hoc test this difference was driven by the overweight versus the obese group (P-value = 0.03945). However, in contrast to this result we found that the B/F ratio, using the Z-score normalization, was not significant different between obese and non-obese individuals (P-value = 0.4516) [Figure 5C]. Additionally, we did not observe a difference when the individuals were separated by BMI groups (P-value = 0.3317) [Figure 5D].

Using a small (0.2), medium (0.5), and large (0.8) cohen value we estimated the power of the combined data set as well as the average cohen value from across studies for Shannon Diversity (0.06) and BF ratio (0.01).

small effect for combined data power calculation:
`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.2, sig.level = 0.05, alternative="two.sided")$power,2), nsmall=2)`  
medium effect for combined data power calculation:
`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.5, sig.level = 0.05, alternative="two.sided")$power,2), nsmall=2)`  
large effect for combined data power calculation:
`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.8, sig.level = 0.05, alternative="two.sided")$power,2), nsmall=2)`  
average effect for all Shannon Diversity:
`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.06, sig.level = 0.05, alternative="two.sided")$power,2), nsmall=2)`  
average effect for all BF ratio:
`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.01, sig.level = 0.05, alternative="two.sided")$power,2), nsmall=2)`

@@

However, other methods do exist, such as body adiposity index (BAI) but do not always correlate with important molecular measures traditionally associated with obesity [@lichtash_body_2013].

* hyper variation between humans may mute the actual signal.

* just because we can't find a human-wide signal, doesn't mean microbiome isn't important for weight

* [@ley_microbial_2006] watched humans lose weight and communities change
* [@turnbaugh_effect_2009] did fecal transplants (humans to mice) and transferred phenotype
* [@koren_host_2012] did fecal transplants (pregnant women to mice) and transferred phenotype

Our meta-analysis helps to provide a little more clarity to the on going debate of whether or not there are specific microbiome variables that are influenced by obesity. Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that the data do not support that there is a microbiome difference between non-obese and obese individuals. However, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give. Here we perform an extensive literature review of the existing studies on the microbiome and obesity and perform a meta-analysis on the studies that remained based on our inclusion and exclusion criteria. We analyze the data using three different approaches. The first approach uses the method taken by the previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014], the second approach takes a meta-analysis approach using RR representation by forest plots and funnel plots, the final approach also takes a meta-analysis approach but uses Z-score normalizations and does not use RR.

Overall, the first approach in this study by and large agrees with the prevous two reviews[@walters_meta-analyses_2014; @finucane_taxonomic_2014], in that for the majority of studies, no significant correlation between the microbiome and obestiy could be observed. However, we found that the other two approaches, that used the pooled data, was able to identify a correlation between measures of the microbiome and obesity [Figure 2 & 3]. For Shannon diversity both approaches using the pooled data found a correlation while only one of the approaches for the B/F ratio found a correlation. Taken together we believe that this analysis provides a strong preliminary finding that there is a small but detectable difference between non-obese and obese individuals with respect to both Shannon diversity and the B/F ratio. However, it should be noted that our analysis would indicate that the signal for Shannon diversity may be more robust than that of the B/F ratio since the two different pooled analysis approaches found Shannon diversity significant while only one did for the B/F ratio.

@@

Second, although we make attempts to normalize the data for the pooled analysis so that we can compare the microbiome across studies it is possible that the different variable regions sampled, sequencing machines used, and the quality of the output data could have an impact on the overall conclusions reached. However, what our analysis shows is that despite these differences, we are still able to identify an overall signal from the data.
Third, this meta-analysis and review of the data to date is still rather preliminary. Only a total of 1542 individuals were included in the pooled analysis which is relatively small compared to other meta-analysis in non-microbiome fields of study. It will be necessary to continuously update this analysis, as more studies become available, in order to obtain a more accurate estimate of the true effect that obesity has on the microbiome within humans.

Fourth, although a significant result exists for the RR for both Shannon diversity and the B/F ratio, it is small. This is also the case for the Z-score normalization analysis of the Shannon diversity. As an example, a lower than median Shannon diversity only has an overall 18% increase of obesity in it versus the higher than median Shannon diversity group. So even though there is a detectable signal it remains to be seen if this is a biologically significant phenomenon that could have a tangible effect on obesity.

Despite the many problems that could add more noise to the measurements used, this study provides a reasonable first pass attempt to pool this data together in an effort to find a correlation between the microbiome and obesity. In doing so we believe that it provides a tanatalizing result that despite the large differences between studies it is still possible to detect a correlation between the microbiome and obesity. However, we emphaszie that it should still be viewed with caution since there are still many aspects of this analysis that could be improved upon. It truly is only a first attempt at trying to summerize and analyze the pooled data that is available. We hope that this study helps to add important information and clarity to the on-going dialogue of the microbiome's role in obesity.

@@

The authors would like to thank Nielson Baxter for his help and suggestions during the sequence analysis portion. We would also like to thank Shawn Whitefield for her help with making sure the manuscript adhered to the guidelines set out for a meta-analysis. Finally, we would like to thank Rick Bushman, Hongzhe Li, and Pixu Shi in agreeing to release and helping to get a hold of the subject specific metadata from the Wu et al study.

@@

```{r Table_5, echo=FALSE, message=FALSE, warning=FALSE}

source("code/MakeFiguresCode.R")
grid.arrange(a2, b2, nrow=2)

```
**Figure 5. Summary of Power and Sample Size Simulations for Shannon Diversity for Non-obese versus Obese**

```{r Table_6, echo=FALSE, message=FALSE, warning=FALSE}

```
**Figure 6. Summary of Power and Sample Size Simulations for Bacteroidetes/Firmicutes Ratio for Non-obese versus Obese**

@@

```{r Figure_2, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}

StudyNo <- as.numeric(ShannonRRTable$X)
Study <- as.character(ShannonRRTable$Study)
Year <- c("2015", "2015", "2014", "2014", "2012", "2011", "2012", "2009")
Total <- c("n=172", "n=63", "n=507", "n=30", "n=200", "n=256", "n=63", "n=146")
tpos <- as.numeric(ShannonRRTable$tposH) #obese in low diversity group
tneg <- as.numeric(ShannonRRTable$tnegH) #nonobese in low diversity group
cpos <- as.numeric(ShannonRRTable$cposH) #obese in high diversity group
cneg <- as.numeric(ShannonRRTable$cnegH) #nonobese in high diversity group
RR <- as.numeric(ShannonRRTable$RRH)
low <- as.numeric(ShannonRRTable$lowH)
high <- as.numeric(ShannonRRTable$highH)

numeric.data <-as.data.frame(cbind(tpos, tneg, cpos, cneg))
raw.overall <- as.data.frame(cbind(StudyNo, Study,Year, Total, numeric.data))

dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data=raw.overall, append=TRUE)
k <- length(dat$StudyNo)
dat.fm <- data.frame(study=factor(rep(1:k, each = 4)))
dat.fm$grp <- factor(rep(c("T", "T", "C", "C"), k), levels = c("T", "C"))
dat.fm$out <- factor(rep(c("+", "-", "+", "-"), k), levels = c("+", "-"))
dat.fm$freq <- with(raw.overall, c(rbind(tpos, tneg, cpos, cneg)))

#Used to generate the RE Model estimate based on all the studies
res <- rma(ai=tpos, bi=tneg, ci=cpos, di=cneg, data=raw.overall, measure="RR", slab=paste(Study, Year, Total, sep=", "), method="REML")
RRSDPvalue <- anova.rma(res)

#To create a graph visualize the data
forest(res, at=log(c(0.3, 1, 5)), atransf=exp, mlab= "Pooled Relative Risk", xlab="RR of Obesity in high versus low Shannon diversity based on Median", cex=0.8)
par(font=2) #Switch to bold

text(5.6, 10.3, "Relative Risk [95% CI]", pos=2, cex=0.8)
text(-6.5, 10.3, "Study, Year, and n", pos=4, cex=0.8)

@@

**Figure 2: Meta Analysis of the Relative Risk of Obesity Based on Shannon Diversity.**  Groups were divided for each study on high and low Shannon Diversity groups based on the median for that study. The overall pooled relative risk was 1.16 for the low diversity group (P-value = `r signif(RRSDPvalue$QMp, 3)`).

@@

StudyNo <- as.numeric(BFRatioRRTable$X)
Study <- as.character(BFRatioRRTable$Study)
Year <- c("2015", "2015", "2014", "2014", "2012", "2011", "2012", "2009")
Total <- c("n=172", "n=63", "n=507", "n=30", "n=200", "n=256", "n=63", "n=146")
tpos <- as.numeric(BFRatioRRTable$tposBF) #obese in low diversity group
tneg <- as.numeric(BFRatioRRTable$tnegBF) #nonobese in low diversity group
cpos <- as.numeric(BFRatioRRTable$cposBF) #obese in high diversity group
cneg <- as.numeric(BFRatioRRTable$cnegBF) #nonobese in high diversity group
RR <- as.numeric(BFRatioRRTable$RRBF)
low <- as.numeric(BFRatioRRTable$lowBF)
high <- as.numeric(BFRatioRRTable$highBF)

numeric.data <-as.data.frame(cbind(tpos, tneg, cpos, cneg))
raw.overall <- as.data.frame(cbind(StudyNo, Study,Year, Total, numeric.data))

dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data=raw.overall, append=TRUE)
k <- length(dat$StudyNo)
dat.fm <- data.frame(study=factor(rep(1:k, each = 4)))
dat.fm$grp <- factor(rep(c("T", "T", "C", "C"), k), levels = c("T", "C"))
dat.fm$out <- factor(rep(c("+", "-", "+", "-"), k), levels = c("+", "-"))
dat.fm$freq <- with(raw.overall, c(rbind(tpos, tneg, cpos, cneg)))

#Used to generate the RE Model estimate based on all the studies
res2 <- rma(ai=tpos, bi=tneg, ci=cpos, di=cneg, data=raw.overall, measure="RR", slab=paste(Study, Year, Total, sep=", "), method="REML")
RRBFPvalue <- anova.rma(res2)

>>>>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load specific R files to source data from for graphs

source("code/MakeRRFiguresCode.R")
source("code/MakeFiguresCode.R")

@@

**Results:** When we considered each dataset in isolation, there were no significant differences (P > 0.05) in B/F between non-obese and obese individuals. Only one diversity measure (OTU richness (N=1)) was significantly different between non-obese and obese individuals (P<0.05). With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals (mean AUC: `r format(round(mean(AUCRFData$OOBAUCAll), 2), nsmall=2)`; s.d.=`r format(round(sd(AUCRFData$OOBAUCAll), 2), nsmall=2)`) and there was no significant difference between the studies (bonferroni adjusted P-value > 0.05). In 7/8 of the studies an OTU related to the Ruminococcaceae family was important in classification; however, the direction of the association varied between studies. When we pooled the results across studies, there was no significant difference in RR of obesity based on Shannon diversity (RR = `r format(round(10^test3R$LogFC[1],2), nsmall=2)`, CI = `r format(round(10^ShannonRes$ci.lb, 2), nsmall=2)`-`r format(round(10^ShannonRes$ci.ub, 2), nsmall=2)`, P-value = `r format(round(RRSDPvalue$QMp, 2), nsmall=2)`) and B/F ratio (RR = `r format(round(10^test3BF$LogFC[1],2), nsmall=2)`, CI = `r format(round(10^BFRes$ci.lb, 2), nsmall=2)`-`r format(round(10^BFRes$ci.ub, 2), nsmall=2)`, P-value = `r format(round(RRBFPvalue$QMp, 2), nsmall=2)`). When we normalized the Shannon diversity data and pooled studies there was also no significant difference between non-obese and obese individuals (difference in mean z-score = `r format(round(combZHTtest$estimate[1]-combZHTtest$estimate[2], 2), nsmall=2)`; P = `r format(round(combZHTtest$p.value, 2), nsmall=2)`). There was also no significant difference in the normalized B/F ratio between non-obese and obese individuals (difference in mean z-score = `r format(round(combZLogBFTtest$estimate[1]-combZLogBFTtest$estimate[2], 2), nsmall=2)`; P = `r format(round(combZLogBFTtest$p.value, 2), nsmall=2)`).  Finally, when completing both power and sample size calculations and simulations we found that most of the studies examined were not sufficiently powered to detect differences in any of the measurements used.

**Conclusion:** We show that by pooling the results of many individual studies we can test the generalizability of the microbiome. Our meta-analysis provides support for the null hypothesis that there is no difference between non-obese and obese individuals in Shannon diversity and B/F ratio, that the effect sizes are small, and that a very large number of indivdiuals would have to be tested to reach sufficient study power.

@@

***Data Analysis***
We performed the analysis using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`). We split the overall analysis into three general strategies. In the first strategy we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B/F), Shannon diversity, and number of OTUs. A two-tailed Wilcoxon test was performed for comparison for non-obese and obese individuals. We also measured the Spearman correlation coefficient between each metric and the actual BMI value. Next, we compared the community structure from non-obese and obese individuals using PERMANOVA analysis of Bray-Curtis distance matrices. This analysis was performed using the vegan R package. Finally, we used the AUCRF (v1.1) R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguish non-obese and obese individuals.  To generate Receiver Operater Characteristic (ROC) curves we used the pROC (v1.8) R package.  A bonferroni adujusted p-value was used to assess differences between each study.

@@

In the third strategy, we normalized the diversity and B/F data within each study and pooled the normalized data across studies. We applied a Z-score normalization to the Shannon diversity data and log-transformed B/F data within each study to have a mean of zero and standard deviation of one. We then pooled Z-scores across studies and used a two-tailed T-test on the values to test for a significant difference in Z-scores among non-obese and obese individuals.

In the final component of the analysis we performed power and sample number simulations for different effect sizes for each study.  This was completed for all the different components measured as part of first strategy.  We also calculated the actual sample size needed based on the effect size of each indivdiual study.  To complete this analysis we used a combination of the pwr (v1.1-3) R package and base R functions.

@@

***Separate Analysis***
We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicute (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014]. Using a P-value cutoff for significance as less than 0.05 we found a only 1 significant result for OTU richness. There were no significant results for Bacteroidetes, Firmicutes, or the B/F ratio. The full summary of the exact P-values for each respective study and variable tested can be found in table S1 of the online supplement.

Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm. We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(AUCRFData$OOBAUCAll), 2), nsmall=2)` +/- `r format(round(sd(AUCRFData$OOBAUCAll), 2), nsmall=2)`. We found that the range for the total number of variables used for the classification ranged from 6 (HMP study) to 42 (Zupancic). The total number of trees and node size was standardized to 1000 and 20 respectively for each data set. Overall there was no significant difference between the ROC curves of each data set (P-value > 0.05, Figure 2).  One aspect we found surprising was that there was very little overlap in the variables used by the random forest algorithm for the classification of obese and non-obese individuals by study. The three most common variables was for an OTU that could be classified to the family Ruminococcaceae (7/8 studies), an OTU that could be classified to the family Lachnospiraceae (6/8 studies), Clostridiales (6/8), Bacteroides (5/8), and Shannon diversity metric (4/8 studies).

Using the top three common taxonomic classifications identified across studies we investigated this further.  Although these taxonomic level classifications showed up commonly across studies, the overall direction and magnitude varied quite a bit depending on the study [Figure S1-3].  Further, when the sequences were BLAST the identifications within each OTU within and between groups varied quite widely with many having relatively poor query coverage and identity [Table S2].  When the sequence identity for each specific variable region was compared using the EMBOSS Needle pairwise sequence alignment (http://www.ebi.ac.uk/Tools/psa/emboss_needle/nucleotide.html) the V4 region had the highest similarity [Table S3] but all had greater than 4% dissimilarity.

@@

We next moved on to the pooled meta-analysis. For the Shannon diversity RR, all studies but Wu, Goodrich, and Turnbaugh had those with a lower Shannon diversity then the median as being higher risk for obesity [Figure 3A]. The was  pooled analysis showed an increased RR of `r format(round(10^test3R$LogFC[1],2), nsmall=2)` (CI `r format(round(10^ShannonRes$ci.lb, 2), nsmall=2)`, `r format(round(10^ShannonRes$ci.ub, 2), nsmall=2)`) for obesity in the lower then median Shannon diversity group that was not significant (P-value = `r format(round(RRSDPvalue$QMp, 2), nsmall=2)`). Similarily, when we investigated the B/F ratio RR  most studies showed that there was no significant difference between obese and non-obese individuals (P-value = `r format(round(RRBFPvalue$QMp, 2), nsmall=2)`)[Figure 3B].  The pooled RR was `r format(round(10^test3BF$LogFC[1],2), nsmall=2)` (CI `r format(round(10^BFRes$ci.lb, 2), nsmall=2)`, `r format(round(10^BFRes$ci.ub, 2), nsmall=2)`).  The differences in RR that was observed also do not seem to be related to the variable region chosen [Figure 3].

We also used a funnel plot to asses the bias of the 9 data sets used in this meta-analysis. From the graph we can observe that there does not seem to be any significant  bias [Figure S4]. Both funnel plots have almost equal number of studies scattered on either side of the estimated true value.

@@

Using the third approach with Z-score normalization we found that there was no significant difference in the Shannon diversity between the obese and non-obese group (P-value = `r format(round(combZHTtest$p.value, 2), nsmall=2)`) [Figure 5A]. Additionally,  we found that the there was also no significant difference in B/F ratio, using the Z-score normalization, between obese and non-obese individuals (P-value = `r format(round(combZLogBFTtest$p.value, 2), nsmall=2)`) [Figure 5B].

**Power and Sample Size Estimates and Simulations**

For Shannon Diversity the measured difference between both obese and non-obese groups was relatively small [Figure 5A] compared to the B/F ratio [Figure 6A] differences.  When simulating different effect size differences of 1%, 5%, 10%, and 15% we found that many studies would be powered sufficiently to find a 15% difference in Shannon diversity [Figure 5A].  However, this was not the case for the B/F ratio where even a 15% difference between non-obese and obese groups only resulted in an average power between studies of `r format(round(mean(ModelledPower15PerDiff$BFratio)*100, 2), nsmall=2)`% [Figure 6A].   Looking at the sample sizes showed that on average to detect a 5% difference at 80% power between non-obese and obese individuals a total n of `r format(round(mean(RRSimNNeededH$RRFivePercentNH), 0), nsmall=0)` would be needed for each group [Figure 5B].  In contrast in order to detect a 5% difference with 80% power in the B/F ratio a sample size of `r format(round(mean(RRSimNNeededBF$RRFivePercentNBF), 0), nsmall=0)` per group [Figure 6B].  A breakdown of the power simulations and sample size estimates of the other variables and RR can be found in the online supplement.

We also investigated if the combined total n for the third strategy was sufficiently powered.  Using a small (0.2), medium (0.5), and large (0.8) cohen value we estimated the power of the combined data sets as well as the average cohen value from across studies for Shannon Diversity (0.06) and BF ratio (0.01).  For the small effect for combined data power calculation a value of `r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.2, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.  For the medium effect for combined data power calculation a value of `r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.5, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.  For the large effect for combined data power calculation a value of
`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.8, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.  The average actual cohens d value for Shannon Diversity for all 8 studies of 0.06 gave a power of `r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.06, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`%.  Finally, the average actual cohens d value for B/F ratio for all 8 studies of 0.01 yieled a power of `r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.01, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`%.

@@

Our meta-analysis helps to provide a little more clarity to the on going debate of whether or not there are specific microbiome variables that are influenced by obesity. Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that the data do not support that there is a microbiome difference between non-obese and obese individuals. However, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give. Here we perform an extensive literature review of the existing studies on the microbiome and obesity and perform a meta-analysis on the studies that remained based on our inclusion and exclusion criteria. We analyzed the data using three different approaches and also performed post-hoc power and sample size calculations and simulations. The first approach uses the method taken by the previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014], the second approach takes a meta-analysis approach using RR representation by forest plots and funnel plots, the final approach also takes a meta-analysis approach but uses Z-score normalizations and does not use RR.

Overall, the first approach in this study by and large agrees with the prevous two reviews[@walters_meta-analyses_2014; @finucane_taxonomic_2014], in that for the majority of studies, no significant correlation between the microbiome and obesityy could be observed. Further, even though random forest could identify specific variables with similar taxonomic classifications that resulted in good ROC curves [Figure 1] these variables varied widely by both their actual sequence and directionality.  Similarily, we found that the other two approaches, that used the pooled data approach, was also not able to identify a correlation between measures of the microbiome and obesity [Figure 3 & 4]. For these approaches both Shannon diversity and B/F ratio failed to have a significant correlation to obesity. In order to also assess the amount of our results that could be due to insufficient power we performed power and sample size analysis.  Overall we found that the most studies for Shannon Diversity would be powered to find a 15% difference, however, the vast majority of reported differences between studies ranged from `r format(round(min(ActualDiff$H), 2), nsmall=2)`-`r format(round(max(ActualDiff$H), 2), nsmall=2)` percent.  Although the effect sizes for the BF ratio were larger than that for Shannon Diversity [Figure 5A & 6A] the estimated power remained low.  Additionaly, we would also need a relatively large sample size to be able to detect the actual differences in each study for both Shannon Diversity and B/F ratio [Figure 5B & 6B].  Taken together we believe that this analysis builds on the previous two re-anlaysis and provides additional information as to whether there is a difference between non-obese and obese human individuals with respect to both Shannon diversity and the B/F ratio.

@@

Second, although we make attempts to normalize the data for the pooled analysis so that we can compare the microbiome across studies it is possible that the different variable regions sampled, sequencing machines used, and the quality of the output data could have an impact on the overall conclusions reached.  To help mitigate some of these problems we analyzed each data set separately and processed all samples through a uniform pipeline in an attempt to mitigate some of the variation that might occur due to sequence processing.  It should also be noted that we also do not see any specific bias in the forest plots [Figure 3] based on hypervariable region.  Further many of the studies show very similar actual effect sizes for Shannon Diversity difference between non-obese and obese individuals.  However, even with all these precautions and safety checks we do realize that it may not be possible to compare these studies together and that further studies need to be completed to verify the observations made in this report.

Third, although we may not have detected a signal between non-obese and obese indivdiuals our results indicate that we may not have been sufficiently powered to do so. However, taking both the sample size calculations and power calculations together would indicate that if there is a true difference to be found between non-obese and obese individuals this difference would be very small.  Additionally, due to the very small signal another important question would be whether or not this difference is truly biologically meaningful.

Fourth, although the ROC curves for each study were relatively good there was very little actual overlap between studies based on sequence similarity and directionality.  This would suggest that within each study classification of obese and non-obese is relatively strong but not generalizable across studies.  This might indicate an ancestry, ethnicity, or a specific environmental effect such as diet difference between study populations that might cause this dissimilarity.  It is also possible that the random forest model overfit the data making it less generalizable across studies.

Although we failed to find an effect it is not realistic to necessarily state that there is no microbiome impact on obesity.  There is a large body of very strong evidence in mice on the effect the microbiome has on obesity [@turnbaugh_effect_2009; @koren_host_2012].  It is not possible to discount these findings and it may be the case that in humans there is just too many other variables and factors that effect the microbiome which makes the signal to hard to detect in a cross sectional manner.  Some well designed studies in humans and diet, in the pasat, have shown that the noise inherent to microbiome studies can be overcome with very careful attention made to controlling the study conditions [@ley_microbial_2006].

Overall, this study provides a reasonable follow up to the two previous re-analysis studies to assess the correlation between the microbiome and obesity. In doing so we believe that we have provided evidence that if there is an effect size present it is small and will take many more studies and samples to be analyzed to confirm a difference between non-obese and obese individuals.

@@

The authors would like to thank Nielson Baxter for his help and suggestions during the sequence analysis portion. We would also like to thank Shawn Whitefield for her help with making sure the manuscript adhered to the guidelines set out for a meta-analysis. We would also like to thank Rick Bushman, Hongzhe Li, and Pixu Shi in agreeing to release and helping to get a hold of the subject specific metadata from the Wu et al study.  Finally, we would like to thank all the authors who have made there data publicly available.  Without this initial forethought studies such as these would not be possible.

@@

@@

```{r Figure_2, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
combinedData <- read.csv("results/tables/ROCcombinedData.csv", stringsAsFactors = FALSE)

diag <- data.frame(x=seq(1, 0), y = seq(0, 1))
diag$Study <- "Control"

rocPlot <- ggplot(data=combinedData, 
                  aes(x = Sensitivity, y = Specificity))
rocPlot + geom_line(aes(group=Study, color=Study)) + 
  scale_colour_brewer(palette="Set1") + 
  scale_x_continuous(trans = "reverse") + 
  geom_line(data=diag, aes(x=x, y=y), linetype="dashed") + 
  theme(axis.line=element_line(colour="black"),
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(),
       panel.border = element_blank(),
       panel.background = element_blank(), 
       legend.key = element_rect(fill = "white"))

@@

**Figure 2: Receiver Operating Characteristic Curve of the Random Forest Analysis of Each Study.** After adjusting for multiple comparisons there was no significant difference between any of the 8 studies examined (P-value > 0.05).

@@

grid.arrange(HRRG1, BFRRG1, ncol=2)

**Figure 3: Meta Analysis of the Relative Risk of Obesity Based on Shannon Diversity or Bacteroidetes/Firmicutes Ratio.**  Groups were divided for each study on high and low Shannon Diversity or Bacteroidetes/Firmicutes (B/F) ratio groups based on the median for that study. Red represents studies that used the 16S rRNA V4 region, yellow represents studies that used a V1-V3 approach, blue represents studies that used either a V1-V2 or just a V2 approach, and orange represents studies that used the V3-V5 region.  **A)** The overall pooled relative risk for Shannon Diversity was `r format(round(10^test3R$LogFC[1],2), nsmall=2)` with a P-value of `r format(round(RRSDPvalue$QMp, 2), nsmall=2)`.  **B)** The overall pooled relative risk for the B/F ratio was  `r format(round(10^test3BF$LogFC[1],2), nsmall=2)` with a P-value of `r format(round(RRBFPvalue$QMp, 2), nsmall=2)`.
